Wrangling multivariate spatio-temporal data with the R package cubble

### 2022-06-13 4-4-netcdf.Rmd
First update of seciton 5.5

The ERA5 dataset [@hersbach2020era5] is the latest reanalysis of global atmosphere, land surface, and ocean waves from 1950 onwards and is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF). The data can be directly downloaded from [Copernicus Climate Data Store (CDS)](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels?tab=overview) website or programmatically via the R package \pkg{ecmwfr} [@ecwmfr]. The \code{era5-pressure} data contains the *specific humidity* and *geopotential* variables on the 10 hPa pressure level on four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04. Once downloaded, the data can be read into a cubble as:

>>>>

The ERA5 dataset [@hersbach2020era5] provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF) and it can be directly downloaded from [Copernicus Climate Data Store (CDS)](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels?tab=overview) website or programmatically via the R package \pkg{ecmwfr} [@ecwmfr]. In this example we focus on the \code{era5-pressure} data which provides hourly data on pressure levels from 1970 to present. Here we specifially look at the information  contained on the variables *specific humidity* measuring the mass of water vapour per kilogram of moist air  and the *geopotential* variable recording the gravitational potential energy of a unit mass, at a particular location, relative to mean sea level. Time-wise we restrict our analysis to four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04. Once downloaded, the data can be read into a \code{cubble} as:

### 2022-06-13 arxiv/cubble-arxiv.tex
First update of seciton 5.5

The ERA5 dataset (\protect\hyperlink{ref-hersbach2020era5}{Hersbach et al. 2020}) is the latest reanalysis of global atmosphere, land surface, and ocean waves from 1950 onwards and is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF). The data can be directly downloaded from \href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels?tab=overview}{Copernicus Climate Data Store (CDS)} website or programmatically via the R package \pkg{ecmwfr} (\protect\hyperlink{ref-ecwmfr}{Hufkens, Stauffer, and Campitelli 2019}). The \code{era5-pressure} data contains the \emph{specific humidity} and \emph{geopotential} variables on the 10 hPa pressure level on four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04. Once downloaded, the data can be read into a cubble as:

>>>>

The ERA5 dataset (\protect\hyperlink{ref-hersbach2020era5}{Hersbach et al. 2020}) provides hourly estimates across the Earth for atmospheric, land and oceanic climate variables. The data is available in the NetCDF format from the European Centre for Medium-Range Weather Forecasts (ECMWF) and it can be directly downloaded from \href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels?tab=overview}{Copernicus Climate Data Store (CDS)} website or programmatically via the R package \pkg{ecmwfr} (\protect\hyperlink{ref-ecwmfr}{Hufkens, Stauffer, and Campitelli 2019}). In this example we focus on the \code{era5-pressure} data which provides hourly data on pressure levels from 1970 to present. Here we specifially look at the information contained on the variables \emph{specific humidity} measuring the mass of water vapour per kilogram of moist air and the \emph{geopotential} variable recording the gravitational potential energy of a unit mass, at a particular location, relative to mean sea level. Time-wise we restrict our analysis to four dates: 2002-09-22, 2002-09-26, 2002-09-30, and 2002-10-04. Once downloaded, the data can be read into a \code{cubble} as:
